Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 numerical-starformation.tex
Removed commented-out text from lots of files - basically vestigial junk.

% \red{( (BWO) I am a little worried that this section is too
%   long. The editorial hatchet may be called for, but I think somebody
%   else (who didn't write the section in the first place) needs to
%   think it through. )}

\subsubsection{Overview}
% Overview of methodology

% \red{(BWO: can this whole paragraph go?)}
% While the physics discussed in other sections of this paper is crucial
% to the study of cosmological structure formation, most observations of
% galaxies in the infrared, optical, and ultraviolet wavelengths are of
% stellar populations.  Furthermore, stars eject energy and metal-
% enriched gas throughout their lives, drastically modifying their own
% environments' cooling properties.  Broadly speaking, the formation of
% galaxies and galaxy clusters cannot be completely modeled without
% including the feedback of energy and metals from stellar populations.
% In particular, it is thought that feedback is crucial for suppressing
% the large numbers of dwarf-like galaxies that CDM theories predict
% \citep{1991ApJ...381...14L,1991ApJ...379...52W}.  An early burst of
% star formation could remove a large fraction of cold gas from such
% systems \citep{1978MNRAS.183..341W,1991ApJ...367...45C}.  Also, the
% unexpectedly low luminosities of small clusters and groups (relative
% to the richest clusters) is commonly explained through feedback
% \citep{1991ApJ...383..104K}.  Energy ejected during the formation of
% the cluster ellipticals lowers the central density and hence the X-ray
% luminosity of such clusters \citep{1997ApJ...484L..21C}.  For all of
% these reasons, it is critical to include star formation and the
% feedback of energy and metals into the interstellar and intergalactic
% media in cosmological simulations.

% Modeling star formation in the context of galaxy formation forces one
% to confront two fundamental problems. The dynamical range required to
% study galaxy formation and evolution is substantial, and attempting to
% directly resolve the formation of individual stars in the context of a
% Milky Way-type galaxy is computationally unfeasible.  Also, a strong
% theoretical understanding of the details of star formation is
% currently lacking.  Due to these challenges, models of star formation
% in cosmological simulations are largely phenomenological, and the
% ``stars'' that are formed are in most cases meant to represent an
% ensemble stellar population (although not all; see
% Section~\ref{sec:starform_pop3}).

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
%The \enzo\ code includes approximately ten models for star formation
%and feedback.  
Broadly speaking, these methods all work in similar
ways: at a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but can also include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, each cell at the
locally highest level of refinement is examined at each timestep to
see if it meets the critiera for star formation.  Star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
%\end{equation} 
%\begin{equation}
\div \myvec{v}_b & < & 0, \\
\label{cencont}
%\end{equation}
%\begin{equation}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
%\end{equation}
%\begin{equation}
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
%$\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
%the mean baryon density in the simulation, 
$\eta$ is the user-defined
overdensity threshold, 
%$\myvec{v}$ is the velocity of the gas in the
%cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
%\rho_{b} + \rho_{d}$, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

% In addition, we have added a stochastic star formation algorithm that
% keeps track of all of the sub-mass threshold stars that should have
% been created and when the total mass of uncreated stars is greater
% than the minimum mass, a star particle is created.

% The feedback of energy and metals into the baryon gas is similar to
% the Kravtsov feedback, with some important differences.

@@

%\begin{eqnarray}  
%\Delta m_{\rm sf} =  m_{*} \left[ \left(1 + \frac{t-t_{\rm
%        form}}{t_{\rm dyn}}  \right)
%%\nonumber \times \\ % uncomment for 2-column
%		   \exp{\left( \frac{-(t-t_{\rm form})}{t_{\rm dyn}} \right) }
%%                    \nonumber \\ 
%	       - \left( 1 + \frac{t+ \Delta t-t_{\rm form}}{t_{\rm dyn}} \right) 
%%	       \nonumber \\ 
%                    \exp{ \left( \frac{-(t+\Delta t-t_{\rm
%                            form})}{t_{\rm dyn}} \right)  }
%                   \right]
%\end{eqnarray}
%which can be represented more clearly in integral form:
%
%\begin{eqnarray}
%\int_{t}^{t+Dt} \frac{dM}{dt} dt = \int_{t}^{t+Dt} m_{*} 
%\frac{dt}{t_{\rm dyn}} \left( \frac{t-t_{\rm form}}{t_{\rm dyn}} \right) 
%%\nonumber \times \\ 
%\exp{ \left( - ~ \frac{t-t_{\rm form}}{t_{\rm dyn}} \right) }
%\end{eqnarray}

@@

The other parameters, such as the overdensity threshold $\eta$,
minimum star mass $m_{\rm *min}$, and mass ejection fraction $f_{m*}$ are
not well constrained either theoretically or observationally.  Indeed,
$m_{\rm *min}$ is a purely numerical parameter designed to keep the code
from producing too many star particles, and thus has no observational
or theoretical counterpart.  The $\eta$ parameter, on the other hand,
nominally has physical meaning (i.e., the density above which star
formation must occur on a very short timescale in a self-gravitating
cloud); however, in the vast majority of simulations the densities
that are reachable are nowhere near the densities of protostellar
clouds, and thus this parameter becomes a rough proxy for finding the
densest environments in a given simulation.  
%Given the
%simulation-dependent nature of these parameters, it is critical to set
%them carefully, and with some sensitivity testing.

@@

This method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
  1959ApJ...129..243S}.  This algorithm is deliberately minimal, and
is explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assumed to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{\rm gas}/\tau_*$. 
%where $\rho_{\rm gas}$ and $\rho_*$ are the
%baryon gas and stellar densities, respectively.  
This ``constant
efficiency'' model on the scale of star formation regions is well
motivated observationally
\citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star formation is
only allowed to take place in very dense regions with $\rho_{\rm gas}
\geq \rho_{\rm SF}$, where $\rho_{\rm SF}$ is a constant proper (as
opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{\rm SF}$ are $\tau_*
= 4$ Gyr and $\rho_{\rm SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_{\rm H}
\sim 50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

@@

% Kravtsov reports that simulations with this algorithm reliably
% reproduce the star formation rate-gas surface density relation of the
% Schmidt law, and are not particularly sensitive to numerical
% parameters \citep{2003ApJ...590L...1K}.  He also notes that this
% algorithm is surprisingly insensitive to the presence or absence of
% feedback and details of the cooling and heating properties of the gas,
% which suggests that the global star formation rate is determined by
% gravitationally- driven supersonic turbulence (on large scales) rather
% than stellar feedback or thermal instabilities on small scales.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ms"
%%% End:

>>>>

\subsubsection{Overview}

Due to the computationally unfeasible number of stars in a galaxy
($10^{11}$) and the lack of detailed understanding of star formation, a
number of phenomenological star formation models are included in
\enzo.
Broadly speaking, these methods all work in similar
ways: at a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but can also include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell with the same
velocity as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, each cell at the
locally highest level of refinement is examined at each timestep to
see if it meets the critiera for star formation.  Star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
\div \myvec{v}_b & < & 0, \\
\label{cencont}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
$\eta$ is the user-defined
overdensity threshold, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

@@

@@

The other parameters, such as the overdensity threshold $\eta$,
minimum star mass $m_{\rm *min}$, and mass ejection fraction $f_{m*}$ are
not well constrained either theoretically or observationally.  Indeed,
$m_{\rm *min}$ is a purely numerical parameter designed to keep the code
from producing too many star particles, and thus has no observational
or theoretical counterpart.  The $\eta$ parameter, on the other hand,
nominally has physical meaning (i.e., the density above which star
formation must occur on a very short timescale in a self-gravitating
cloud); however, in the vast majority of simulations the densities
that are reachable are nowhere near the densities of protostellar
clouds, and thus this parameter becomes a rough proxy for finding the
densest environments in a given simulation.

@@

This method of star particle creation is designed to reproduce
the global Schmidt law of star formation~\citep{2003ApJ...590L...1K,
  1959ApJ...129..243S}.  This algorithm is deliberately minimal, and
is explicitly geared towards modeling star formation in a
phenomenological way on kiloparsec scales.  Stars are assumed to form
with a characteristic gas timescale $\tau_*$ such that $\dot{\rho}_* =
\rho_{\rm gas}/\tau_*$. 
This ``constant
efficiency'' model on the scale of star formation regions is well
motivated observationally
\citep{1996AJ....112.1903Y,2002ApJ...569..157W}.  Star formation is
only allowed to take place in very dense regions with $\rho_{\rm gas}
\geq \rho_{\rm SF}$, where $\rho_{\rm SF}$ is a constant proper (as
opposed to comoving) density threshold.  No other criteria are
imposed.  Typical choices for $\tau_*$ and $\rho_{\rm SF}$ are $\tau_*
= 4$ Gyr and $\rho_{\rm SF} = 1.64~$M$_\odot$~pc$^{-3}$~$(n_{\rm H}
\sim 50$~cm$^{-3})$.  The adopted timescale is derived from the
observationally-determined normalization of the Schmidt law, and the
density threshold is determined by observations of star forming
regions on $\sim 100$ pc scales.

@@

@@

### 2013-06-25 numerical-timestep.tex
Removed commented-out text from lots of files - basically vestigial junk.

% \begin{equation}
% \Delta t_{} = 
% \label{eqn:dt}
% \end{equation}

% \begin{equation}
% \Delta t_{} = 
% \label{eqn:dt}
% \end{equation}

@@

Equation~(\ref{eqn:dthydro}) ensures that all cells satisfy the
Courant-Freidrichs-Levy (CFL) condition for accuracy and stability of
an explicit finite difference discretization of the Euler equations.
%Effectively this condition forces the timestep to be small enough such
%that any changes in the fluid propagate less than a single grid
%spacing, $\Delta x$.  
In this equation, $\kappa_{\rm hydro}$ is a dimensionless
numerical constant with a value of $0 < \kappa_{\rm hydro} \leq 1$
(with a typical value of $\kappa_{\rm hydro} \sim 0.3-0.5$) that
ensures that the CFL condition is always met, and $c_s$ and $v_x$ are
the sound speed and peculiar baryon velocity in a given cell.

@@

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, 
%defined as $n_b = \rho_b / \mu
%m_p$, where $\mu$ is the mean molecular weight, 
$\kappa_{\rm sp}(T)$ is the Spitzer thermal conductivity, and f$_{\rm sp}$ is a
user-defined, dimensionless
conduction suppression factor whose value must be f$_{\rm sp} \leq 1$.  
%Both are defined in Section~\ref{sec.num.conductions}.  
$k_{\rm cond}$ is a dimensionless prefactor whose
value must be $0 < k_{\rm cond}  \leq 0.5$, and is exactly 0.5 for the
implemention in \enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$ with
the grid cell size $\Delta x$ and temperature $T$),
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.

>>>>

@@

Equation~(\ref{eqn:dthydro}) ensures that all cells satisfy the
Courant-Freidrichs-Levy (CFL) condition for accuracy and stability of
an explicit finite difference discretization of the Euler equations.
In this equation, $\kappa_{\rm hydro}$ is a dimensionless
numerical constant with a value of $0 < \kappa_{\rm hydro} \leq 1$
(with a typical value of $\kappa_{\rm hydro} \sim 0.3-0.5$) that
ensures that the CFL condition is always met, and $c_s$ and $v_x$ are
the sound speed and peculiar baryon velocity in a given cell.

@@

Equation~(\ref{eqn:dtcond}) is the stability condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, 
$\kappa_{\rm sp}(T)$ is the Spitzer thermal conductivity, and f$_{\rm sp}$ is a
user-defined, dimensionless
conduction suppression factor whose value must be f$_{\rm sp} \leq 1$.  
%Both are defined in Section~\ref{sec.num.conductions}.  
$k_{\rm cond}$ is a dimensionless prefactor whose
value must be $0 < k_{\rm cond}  \leq 0.5$, and is exactly 0.5 for the
implemention in \enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$ with
the grid cell size $\Delta x$ and temperature $T$),
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.
